---
title: "alpinePhenology_Bayes"
author: "emlombardi"
date: "2024-05-08"
output: html_document
---

# Introduction
This is a script to create and run new analyses for Erin's phenology project. Specifically, these are the models we will work on for the manuscript. The goal is to model flowering date over time for each of the different species with specific relationships (hypothetical) to variables like precipitation, temperature, year, site, etc. 

Resources that may be useful:
https://m-clark.github.io/mixed-models-with-R/bayesian.html
https://mc-stan.org/rstanarm/

```{r libraries}
#Install as needed
install.packages("rstanarm")
install.packages("bayesplot")

#Load libraries
library(ggplot2)
library(bayesplot)
library(rstanarm)
library(lme4)
library(broom.mixed)


#Load datastack
load("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/phenologyCollProj_workspace.RData")

#or

load("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/phenAbio.RData")

```

#Models

##Basic linear models (frequentist)
First, let's build some basic linear models to explore the data

### linear model minimum ordinal date/year for each species
This is the most basic version. Is there are relationship between earliest observation for each species across the history of the sites? 

```{r}

#each species
spp <- unique(phen.abio$scientific_name) #15 unique species in the dataset
lm_models <- list()

# Loop over each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, spp == i)
  
  #aggregate to find minimum date for each species in each year
  min_data <- aggregate(ordinal_date ~ year, data = subset_data, FUN = min)
  
  # Lm for minimum ordinal date each year for each species
  lm_model <- lm(ordinal_date ~ year, data = min_data)
  
  # Store the lm model in the list with species name as key
  lm_models[[i]] <- lm_model
}

summary(lm_models$`Polemonium viscosum`)
summary(lm_models)
str(lm_models)
class(lm_models) #lm_models is a recursive list



##Extract the relevant elements for each species
model_summaries_lm <- data.frame(Species = character(), 
                              Coefficients = character(),
                              Std_Errors = numeric(),
                              t_Values = numeric(),
                              p_Values = numeric(),
                              stringsAsFactors = FALSE)

for (i in spp) {
  # Extract summary information from the linear model
  summary_info <- summary(lm_models[[i]])
  
  # Extract coefficients, standard errors, t-values, and p-values
  coefficients <- summary_info$coefficients[, 1]
  std_errors <- summary_info$coefficients[, 2]
  t_values <- summary_info$coefficients[, 3]
  p_values <- summary_info$coefficients[, 4]
  
  # Create a data frame with the extracted information
  species_summary <- data.frame(Species = i,
                                Coefficients = rownames(summary_info$coefficients),
                                Std_Errors = std_errors,
                                t_Values = t_values,
                                p_Values = p_values,
                                stringsAsFactors = FALSE)
  
  # Append the species summary to the model_summaries data frame
  model_summaries_lm <- rbind(model_summaries_lm, species_summary)
}

print(model_summaries_lm) #this is the basic linear model looking at the minimum ordinal date per species per year over time
write.csv(model_summaries_lm, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_lm.csv")

```


### mixed linear model for species flowering data over time
This is the most basic version. Is there are relationship between earliest observation for each species across the history of the sites? 


##### MIXED FREQ MODELS
```{r}
# Initialize an empty list to store models
mixed_models <- list()
spp <- unique(phen.abio$scientific_name)  


# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, scientific_name == i)
  
  # Find the earliest observation for each year while keeping all variables
  earliest_data <- subset_data %>%
    group_by(year, prevWintPPT, data_type) %>%
    filter(ordinal_date == min(ordinal_date)) %>%
    ungroup()
  
  # Fit mixed-effects model with fixed effects for year and random effects for data_type; when I included site it was over-fit
  mixed_model <- lmer(ordinal_date ~ year + prevWintPPT+ (1 | data_type), data = earliest_data)
  
  # Store the mixed-effects model in the list with species name as key
  mixed_models[[i]] <- mixed_model
}

model_summaries <- lapply(mixed_models, summary)
print(model_summaries$`Primula angustifolia`)


##TIDY model output for all earliest/spp data
# Create a tidy table for all fitted models
model_summaries <- lapply(names(mixed_models), function(species) {
  model <- mixed_models[[species]]
  tidy_model <- tidy(model, effects = "fixed")  # Extract fixed effects
  tidy_model$species <- species  # Add species name to each row
  tidy_model
})

all_summaries <- bind_rows(model_summaries)
print(all_summaries)

write.csv(all_summaries, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_mixedFrequentist_early.csv", row.names = FALSE)


```

```{r}
# Initialize an empty list to store models
mixed_models <- list()
spp <- unique(phen.abio$scientific_name)

i='Eremogone fendleri'
# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, scientific_name == i)
  
  # Find the earliest observation for each year while keeping all variables
  earliest_data <- subset_data %>%
    #group_by(year, prevWintPPT, data_type) %>%
    group_by(year) %>%
    filter(ordinal_date == min(ordinal_date)) %>%
    ungroup()
  
  # Fit mixed-effects model with fixed effects for year and prevWintPPT, and random effects for data_type
  mixed_model <- lmer(ordinal_date ~ year + prevWintPPT + (1 | data_type), data = earliest_data)
  
  # Store the mixed-effects model in the list with species name as key
  mixed_models[[i]] <- mixed_model
}

# Create a tidy table for all fitted models including p-values and t-values
all_summaries <- lapply(names(mixed_models), function(species) {
  model <- mixed_models[[species]]
  
  # Extract fixed effects and calculate p-values and t-values
  tidy_model <- tidy(model, effects = "fixed") %>%
    mutate(
      species = species,
      p.value = 2 * pt(abs(statistic / std.error), df.residual(model), lower.tail = FALSE),
      t.value = statistic / std.error
    ) %>%
    select(species, term, estimate, std.error, statistic, t.value, p.value)
  
  tidy_model
})

# Combine all summaries into a single data frame
all_summaries_df <- bind_rows(all_summaries)

# Print the tidy table
print(all_summaries_df)


write.csv(all_summaries, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_mixedFrequentist_early.csv", row.names = FALSE)

#NOTE TO SELF: I think I might want to rescale the year variable so it starts at 0 and each year is a new integer, or something. 

```

# Bayesian linear model
## Using BLME package

```{r}
# Initialize an empty list to store models
mixed_models <- list()
spp <- unique(phen.abio$scientific_name)

# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, scientific_name == i)
  
  # Find the earliest observation for each year while keeping all variables
  earliest_data <- subset_data %>%
    group_by(year) %>%
    filter(ordinal_date == min(ordinal_date)) %>%
    ungroup()
  
  # Fit Bayesian mixed-effects model with fixed effects for year and prevWintPPT, and random effects for data_type
  mixed_model_b <- blmer(ordinal_date ~ year + prevWintPPT + (1 | data_type), 
                         data = earliest_data,
                         prior=normal(0,10),
                         prior_intercept=normal(0,10))
  
  # Store the mixed-effects model in the list with species name as key
  mixed_models[[i]] <- mixed_model_b
}

# Print one of the fitted models for verification
print(summary(mixed_models$`Primula angustifolia`))

# Optionally, you can create a tidy table with all fitted models including p-values and t-values
model_summaries <- lapply(names(mixed_models), function(species) {
  model <- mixed_models[[species]]
  
  # Extract fixed effects and other relevant information
  tidy_model <- tidy(model, effects = "fixed") %>%
    mutate(
      species = species
    ) %>%
    select(species, term, estimate, std.error, statistic, p.value)
  
  tidy_model
})

# Combine all summaries into a single data frame
all_summaries <- bind_rows(model_summaries)

# Print the tidy table
print(all_summaries)



```




# BRMSpackage: Mixed model for earliest flowering dates per species over time



###BRMS model winter ppt 
This first chunk is the basic model with just preceding-winter temperature as a variable. I'll also try preceding ppt and current season temperature, and might choose to re-work precipiation so it's not the mean of monthly means, but rather the total winter precipitation. I'll do that after Botany, though!

```{r}

# Install and load necessary libraries
if (!requireNamespace("brms", quietly = TRUE)) {
  install.packages("brms")
}
library(brms)
library(dplyr)

# Initialize an empty list to store models
bayesian_models <- list()
spp <- unique(phen.abio$scientific_name)


#for testing
#i='Cherleria obtusiloba'

# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, scientific_name == i)
  
  # Find the earliest observation for each year while keeping all variables
  earliest_data <- subset_data %>%
    group_by(year) %>%
    filter(ordinal_date == min(ordinal_date)) %>%
    ungroup() 
  
  # Fit Bayesian mixed-effects model with fixed effects for year and random effects for data_type and site
  bayesian_model <- brm(
    ordinal_date ~ year + prevWintPPT + data_type,
    data = earliest_data,
    chains = 4,  # Number of Markov chains
    iter = 2000, # Number of iterations per chain
    warmup = 1000, # Number of warmup (burn-in) iterations
    cores = parallel::detectCores(), # Use all available cores
    control = list(adapt_delta = 0.95, max_treedepth = 15)
    )
  
  # Store the Bayesian model in the list with species name as key
  bayesian_models[[i]] <- bayesian_model
}


# Function to extract tidy model summaries and add significance
extract_model_summary <- function(model, species_name) {
  summary_df <- tidy(model, effects = "fixed", conf.int = TRUE)
  summary_df$species <- species_name
  summary_df <- summary_df %>%
    mutate(significant = ifelse(conf.low > 0 | conf.high < 0, TRUE, FALSE))
  return(summary_df)
}


# Extract summaries for all models and add species names
model_summaries <- map2_dfr(bayesian_models, names(bayesian_models), extract_model_summary)


#Save all model outputs; note that the data_type is not currently working, I don't think...probably need to define preiors better
write.csv(model_summaries, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_Bayesian.csv", row.names = FALSE)



#extract only those results with CI intervals that indicate significance
# Extract only significant results
significant_results <- model_summaries %>%
  filter(significant == TRUE)

print(significant_results) #this is probably what I want to report on!

write.csv(significant_results, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_Bayesian_SigIndividSPP.csv", row.names = FALSE)



###PLOT the results
# Function to plot fixed effects for a given model
plot_fixed_effects <- function(model, species_name) {
  fixed_effects <- tidy(model, effects = "fixed")
  ggplot(fixed_effects, aes(x = term, y = estimate, ymin = estimate - std.error, ymax = estimate + std.error)) +
    geom_pointrange() +
    labs(title = paste("Fixed Effects for", species_name), x = "Term", y = "Estimate") +
    theme_minimal()
}

# Generate plots for all species
plots <- map2(bayesian_models, names(bayesian_models), plot_fixed_effects)


# Combine all plots using patchwork
library(patchwork)
combined_plot <- wrap_plots(plots, ncol = 3)
combined_plot


ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Figures/combined_plots_Bayesian.png", combined_plot, width = 24, height = 18)



```


##BRMS model with temperature too
```{r}

# Initialize an empty list to store models
bayesian_models <- list()
spp <- unique(phen.abio$scientific_name)


#for testing
#i='Cherleria obtusiloba'

# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen.abio, scientific_name == i)
  
  # Find the earliest observation for each year while keeping all variables
  earliest_data <- subset_data %>%
    group_by(year) %>%
    filter(ordinal_date == min(ordinal_date)) %>%
    ungroup() 
  
  # Fit Bayesian mixed-effects model with fixed effects for year and random effects for data_type and site
  bayesian_model <- brm(
    ordinal_date ~ year + prevWintPPT + tmean.avg + data_type,
    data = earliest_data,
    chains = 4,  # Number of Markov chains
    iter = 2000, # Number of iterations per chain
    warmup = 1000, # Number of warmup (burn-in) iterations
    cores = parallel::detectCores(), # Use all available cores
    control = list(adapt_delta = 0.95, max_treedepth = 15)
    )
  
  # Store the Bayesian model in the list with species name as key
  bayesian_models[[i]] <- bayesian_model
}


# Function to extract tidy model summaries and add significance
extract_model_summary <- function(model, species_name) {
  summary_df <- tidy(model, effects = "fixed", conf.int = TRUE)
  summary_df$species <- species_name
  summary_df <- summary_df %>%
    mutate(significant = ifelse(conf.low > 0 | conf.high < 0, TRUE, FALSE))
  return(summary_df)
}


# Extract summaries for all models and add species names
model_summaries <- map2_dfr(bayesian_models, names(bayesian_models), extract_model_summary)


#Save all model outputs; note that the data_type is not currently working, I don't think...probably need to define preiors better
write.csv(model_summaries, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_Bayesian_TempPPTboth.csv", row.names = FALSE)



#extract only those results with CI intervals taht indicate significance
# Extract only significant results
significant_results <- model_summaries %>%
  filter(significant == TRUE)

print(significant_results) #this is probably what I want to report on!

write.csv(significant_results, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_Bayesian_SigIndividSPP_TempPPTboth.csv", row.names = FALSE)



###PLOT the results
# Function to plot fixed effects for a given model
plot_fixed_effects <- function(model, species_name) {
  fixed_effects <- tidy(model, effects = "fixed")
  ggplot(fixed_effects, aes(x = term, y = estimate, ymin = estimate - std.error, ymax = estimate + std.error)) +
    geom_pointrange() +
    labs(title = paste("Fixed Effects for", species_name), x = "Term", y = "Estimate") +
    theme_minimal()
}

# Generate plots for all species
plots <- map2(bayesian_models, names(bayesian_models), plot_fixed_effects)


# Combine all plots using patchwork
library(patchwork)
combined_plot <- wrap_plots(plots, ncol = 3)
combined_plot


ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Figures/combined_plots_BayesianTempPPT.png", combined_plot, width = 24, height = 18)

png("pairs_plot.png", width = 2000, height = 2000)
plot(bayesian_model)
dev.off()


```


### Mixed model for median flowering dates per species over time

```{r}
# Initialize an empty list to store models
bayesian_models_median <- list()

# Loop through each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen, spp == i)
  
  # Find the earliest observation for each year while keeping all variables
  # median_data <- subset_data %>%
  #   group_by(year) %>%
  #   filter(ordinal_date == median(ordinal_date)) %>%
  #   ungroup()
  # 
  # Fit Bayesian mixed-effects model with fixed effects for year and random effects for data_type and site
  bayesian_model <- brm(
    ordinal_date ~ year + (1 | data_type) + (1 | site),
    data = subset_data,
    chains = 4,  # Number of Markov chains
    iter = 2000, # Number of iterations per chain
    warmup = 1000, # Number of warmup (burn-in) iterations
    cores = parallel::detectCores() # Use all available cores
  )
  
  # Store the Bayesian model in the list with species name as key
  bayesian_models_median[[i]] <- bayesian_model
}

# Extract model summaries using lapply
bayesian_models_median_summaries <- lapply(bayesian_models_median, summary)

# Print the summary of the first Bayesian model as an example
print(bayesian_models_median_summaries[[1]])

# Extract posterior summaries from the first model
posterior_summary_median <- posterior_summary_median(bayesian_models_median[[1]])
print(posterior_summary_median)

plot(posterior_summary_median)
```


# Altered: species as random variable

```{r}

#adjust year data for the brms model (starts from 1871=year 1)
phen <- phen %>%
  mutate(year_mod = I(year-1871)) #maybe scale by adding "center=T" outside if inner parentheses



###FOR ALL DATA

#no priors
m1 <- brms::brm(
  formula = ordinal_date ~ year_mod + (1 | site) + (1 | scientific_name),
  data = phen,
  family = gaussian(),  # Assuming a normal distribution for 'doy'
  iter = 2000,  # Number of iterations
  warmup = 500,  # Number of warm-up iterations
  chains = 4,   # Number of Markov chains
  cores = 4,     # Number of cores for parallel processing
  control = list(max_treedepth=15, adapt_delta = 0.99)  # Increased from default; brings the number of failures down to 15 
)
pairs(m1)
summary(m1)
plot(m1)

#scientific name as random effect
m2 <- brms::brm(
  formula = ordinal_date ~ year_mod + (1 | site) + scientific_name,
  data = phen,
  family = gaussian(),  # Assuming a normal distribution for 'doy'
  iter = 2000,  # Number of iterations
  warmup = 500,  # Number of warm-up iterations
  chains = 4,   # Number of Markov chains
  cores = 4,     # Number of cores for parallel processing
  control = list(max_treedepth=15, adapt_delta = 0.99)  # Increased from default; brings the number of failures down to 15 
)
pairs(m2)
summary(m2)


```





#
##
###
# Phenomix package max likelihood modeling approach
Recent new package described here: https://esajournals.onlinelibrary.wiley.com/doi/10.1002/ecy.4061
GitHub page here: https://github.com/nwfsc-cb/phenomix

update: I don't think this will be helpful for us. The package requires count data (i.e. # of events per day over years), which would only be sensible if we grouped by species. We will have lots of 0 and 1 values. 

```{r}
remotes::install_github("nwfsc-cb/phenomix",build_vignettes = TRUE)

library(phenomix)
library(ggplot2)
library(dplyr)
library(TMB)

```

