---
title: "alpinePhenology_Bayes"
author: "emlombardi"
date: "2024-05-08"
output: html_document
---

# Introduction
This is a script to create and run new analyses for Erin's phenology project. Specifically, these are the models we will work on for the manuscript. The goal is to model flowering date over time for each of the different species with specific relationships (hypothetical) to variables like precipitation, temperature, year, site, etc. 

Resources that may be useful:
https://m-clark.github.io/mixed-models-with-R/bayesian.html
https://mc-stan.org/rstanarm/

```{r libraries}
#Install as needed
install.packages("rstanarm")
install.packages("bayesplot")

#Load libraries
library(ggplot2)
library(bayesplot)
library(rstanarm)
library(lme4)




#Load datastack
load("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/phenologyCollProj_workspace.RData")

```

#Models

##Basic linear models (frequentist)
First, let's build some basic linear models to explore the data

### linear model minimum ordinal date/year for each species
This is the most basic version. Is there are relationship between earliest observation for each species across the history of the sites? 

```{r}

#each species
spp <- unique(phen$scientific_name) #15 unique species in the dataset
lm_models <- list()

# Loop over each species
for (i in spp) {
  # Subset data for the current species
  subset_data <- subset(phen, spp == i)
  
  #aggregate to find minimum date for each species in each year
  min_data <- aggregate(ordinal_date ~ year, data = subset_data, FUN = min)
  
  # Lm for minimum ordinal date each year for each species
  lm_model <- lm(ordinal_date ~ year, data = min_data)
  
  # Store the lm model in the list with species name as key
  lm_models[[i]] <- lm_model
}

summary(lm_models$`Polemonium viscosum`)
summary(lm_models)
str(lm_models)
class(lm_models) #lm_models is a recursive list



##Extract the relevant elements for each species
model_summaries_lm <- data.frame(Species = character(), 
                              Coefficients = character(),
                              Std_Errors = numeric(),
                              t_Values = numeric(),
                              p_Values = numeric(),
                              stringsAsFactors = FALSE)

for (i in spp) {
  # Extract summary information from the linear model
  summary_info <- summary(lm_models[[i]])
  
  # Extract coefficients, standard errors, t-values, and p-values
  coefficients <- summary_info$coefficients[, 1]
  std_errors <- summary_info$coefficients[, 2]
  t_values <- summary_info$coefficients[, 3]
  p_values <- summary_info$coefficients[, 4]
  
  # Create a data frame with the extracted information
  species_summary <- data.frame(Species = i,
                                Coefficients = rownames(summary_info$coefficients),
                                Std_Errors = std_errors,
                                t_Values = t_values,
                                p_Values = p_values,
                                stringsAsFactors = FALSE)
  
  # Append the species summary to the model_summaries data frame
  model_summaries_lm <- rbind(model_summaries_lm, species_summary)
}

print(model_summaries_lm) #this is the basic linear model looking at the minimum ordinal date per species per year over time
write.csv(model_summaries_lm, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/model_summaries_lm.csv")

```
