---
title: "PhenologyProj_PhyloDiversity"
author: "emlombardi"
date: "2024-01-17"
output: html_document
---

#Introduction

We have already looked at earliest flowering across the focal species in Erin's analysis and have some results indicating first flowering is advancing across most. We may also want to see if there is a phylogenetic signal that we may detect. 

To do this, I will be referring to two online resources:
1. Tutorial ![https://pedrohbraga.github.io/CommunityPhylogenetics-Workshop/CommunityPhylogenetics-Workshop.html#within-assemblage-phylogenetic-structure]
2. Hannah's sky island paper/methods ![https://onlinelibrary-wiley-com.libproxy.unm.edu/doi/full/10.1111/jbi.13073]

```{r libraries}
# Check for needed packages, and install the missing ones
required.libraries <- c("ape", "picante", 
                        "pez", "phytools",
                        "vegan", "adephylo", 
                        "phylobase", "geiger", 
                        "mvMORPH", "OUwie", 
                        "hisse", "BAMMtools",
                        "phylosignal", "Biostrings",
                        "devtools","ggplot2", 
                        "kableExtra", "betapart", "gridExtra",
                        "reshape2")

needed.libraries <- required.libraries[!(required.libraries %in% installed.packages()[,"Package"])]
if(length(needed.libraries)) install.packages(needed.libraries)

install.packages("Biostrings")

# Load all required libraries at once
lapply(required.libraries, require, character.only = TRUE)

### Install ggtree from BiocManager

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ggtree")

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
