---
title: "Phenology_TaxonomicChangeTime"
author: "emlombardi"
date: "2024-01-16"
output: html_document
---

#Is there a phylogenetic signal in flowering advancement across time?

This is a script to look at taxonomic and phylogenetic variation in first flowering at Pikes Peak and Mount Blue Sky in Colorado. Created by EML but intended for use by EB.

```{r libraries}
library(dplyr)
library(ggplot2)
library(lubridate)

```
##How many observations and specimens per species?
Note that there are far more observations than there are specimens (N=1371 observations, N=514 specimens).

There are different numbers of data points for each species, and this varies by data type. Nonetheless, it's entirely possible that there are enough across time to assess across time.

```{r}
#observations
observations <- phen2 %>%
  filter(data_type == "observation")
tObs<-observations %>%
  group_by(scientific_name) %>%
  tally(sort=TRUE)%>%
  rename("cnt.Obs"="n")


#specimens
specimen <- phen2 %>%
  filter(data_type == "specimen")
tSpec<-specimen %>%
  group_by(scientific_name) %>%
  tally(sort=TRUE) %>%
  rename("cnt.Specimen"="n")

cntsTable <- tObs %>%
  left_join(tSpec, by="scientific_name")

write.csv(cntsTable, "/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Tables/table_TallybySpecies.csv", row.names=TRUE)

```


##When were different species recorded over time?

```{r}
#observations
recs1<-ggplot(observations, aes(x = year, y = ordinal_date)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "lm", se = FALSE, color="goldenrod") +
  labs(
    title = "Box Plot of Values by Category",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)

ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Figures/SpeciesObservations_OverTime.pdf", recs1, width = 12, height = 8, dpi = 300)


#Take a look differently:
ggplot(observations, aes(x = year)) +
  geom_bar() +
  labs(
    title = "Box Plot of Values by Category",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)



#specimens with slope values
recs2<-ggplot(specimen, aes(x = year, y = ordinal_date)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "lm", se = FALSE, color="skyblue3") +
  labs(
    title = "Box Plot of Values by Category",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)

ggsave("/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Figures/SpeciesSpecimen_OverTime.pdf", recs2, width = 12, height = 8, dpi = 300)

ggplot(specimen, aes(x = year)) +
  geom_bar() +
  labs(
    title = "Box Plot of Values by Category",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)




```
##Changing earliest flowering date for each species?
There aren't enough records made each year for each species to really assess change over time, but maybe we could use at a decadal time unit. 
What I want to do here:
-Calculate the number of specimen records made for each species each decade
-Find the earliest ordinal date for each species each decade
-Have first flowering dates each decade changed?

I think I should just create a new dataframe with columns: scientific_name, Number of records/decade, earliest flowering date/decade

####Specimen data
```{r}
#step 1: new dataframe with earliest flowering date each year for each species
minSpec <- specimen %>%
  group_by(scientific_name, year) %>%
  summarize(minDate = min(ordinal_date))

#add slope info for each species
minSpec<- minSpec%>%
  group_by(scientific_name) %>%
  mutate(slope = coef(glm(minDate ~ year))[2])

ggplot(minSpec, aes(x = year, y = minDate)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "glm", se = FALSE, color="skyblue3") +
  geom_text(data = minSpec %>% slice(1),  # Take the first row for each species
            aes(label = paste("Slope:", round(slope, 3))),
            x = Inf, y = Inf, hjust = 1, vjust = 1,
            color = "black", size = 3) +
  labs(
    title = "Earliest recorded flowering for each species each year",
    x = "Year",
    y = "Ordinal date"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)



####DECADAL change
#Group years into decades and find earliest flowering date for each species each decade
specimen<- specimen %>%
  mutate(decade=floor(year/10)*10) 

#calculate the minimum date
minSpec.Decade <- specimen %>%
  group_by(scientific_name, decade) %>%
  summarize(minDate = min(ordinal_date))

#add slope info for each species
minSpec.Decade <- minSpec.Decade %>%
  group_by(scientific_name) %>%
  mutate(slope = coef(glm(minDate ~ decade))[2])


ggplot(minSpec.Decade, aes(x = decade, y = minDate)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "glm", se = FALSE, color="skyblue3") +
geom_text(data = minSpec.Decade %>% slice(1),  # Take the first row for each species
            aes(label = paste("Slope:", round(slope, 3))),
            x = Inf, y = Inf, hjust = 1, vjust = 1,
            color = "black", size = 3) +
  labs(
    title = "Earliest recorded flowering for each species in each decade",
    x = "Decade",
    y = "Ordinal date"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)

```

###Statistical significance of slopes?

```{r}
library(broom)

####ANNUAL MODELS
mod1 <- glm(minDate ~ year + scientific_name, data = minSpec)
tidymod1 <- as.data.frame(tidy(mod1))
summary(mod1)

pvals.mod1 <- tidymod1 %>%
  filter(!term == "year") %>%
  select(term, p.value)

mod1.sigSpp <- pvals.mod1 %>% #significant decrease at the ANNUAL time scale for a few
  filter(p.value < 0.05)


####DECADAL MODELS
mod2 <- glm(minDate ~ decade + scientific_name, data = minSpec.Decade)
tidymod2 <- tidy(mod2)
summary(mod2)


pvals.mod2 <- tidymod2 %>%
  #filter(term == "decade") %>%
  select(term, p.value)

mod2.sigSpp <- pvals.mod2 %>% #significant decrease at the decadal time scale for a few
  filter(p.value < 0.05)



```



####Observational Data

```{r}
#step 1: new dataframe with earliest flowering date each year for each species
minObs <- observations %>%
  group_by(scientific_name, year) %>%
  summarize(minDate = min(ordinal_date))

#add slope info for each species
minObs<- minObs%>%
  group_by(scientific_name) %>%
  mutate(slope = coef(glm(minDate ~ year))[2])

ggplot(minObs, aes(x = year, y = minDate)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "glm", se = FALSE, color="goldenrod3") +
  geom_text(data = minObs %>% slice(1),  # Take the first row for each species
            aes(label = paste("Slope:", round(slope, 3))),
            x = Inf, y = Inf, hjust = 1, vjust = 1,
            color = "black", size = 3) +
  labs(
    title = "Earliest recorded flowering for each species each year",
    x = "Year",
    y = "Ordinal date"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)



####DECADAL change
#Group years into decades and find earliest flowering date for each species each decade
observations <- observations %>%
  mutate(decade=floor(year/10)*10) 

#calculate the minimum date
minObs.Decade <- observations %>%
  group_by(scientific_name, decade) %>%
  summarize(minDate = min(ordinal_date))

#add slope info for each species
minObs.Decade <- minObs.Decade %>%
  group_by(scientific_name) %>%
  mutate(slope = coef(glm(minDate ~ decade))[2])


ggplot(minObs.Decade, aes(x = decade, y = minDate)) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "glm", se = FALSE, color="goldenrod3") +
geom_text(data = minObs.Decade %>% slice(1),  # Take the first row for each species
            aes(label = paste("Slope:", round(slope, 3))),
            x = Inf, y = Inf, hjust = 1, vjust = 1,
            color = "black", size = 3) +
  labs(
    title = "Earliest recorded flowering for each species in each decade",
    x = "Decade",
    y = "Ordinal date"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)

```

###Statistical significance of slopes?

```{r}
library(broom)

####ANNUAL MODELS
mod1b <- glm(minDate ~ year + scientific_name, data = minObs)
tidymod1b <- as.data.frame(tidy(mod1))
summary(mod1b)

pvals.mod1b <- tidymod1b %>%
  filter(!term == "year") %>%
  select(term, p.value)

mod1b.sigSpp <- pvals.mod1b %>% #significant decrease at the ANNUAL time scale for a few
  filter(p.value < 0.05)



####DECADAL MODELS
mod2b <- glm(minDate ~ decade + scientific_name, data = minObs.Decade)
tidymod2b <- tidy(mod2b)
summary(mod2b)

pvals.mod2b <- tidymod2b %>%
  #filter(term == "decade") %>%
  select(term, p.value)

mod2b.sigSpp <- pvals.mod2b %>% #significant decrease at the decadal time scale for a few
  filter(p.value < 0.05)

```