---
title: 'Frost Index: phenology project'
author: "emlombardi"
date: "2024-10-31"
output: html_document
---

The objectives of this script are to:
1. calculate an index that categorizes years with high, moderate and low spring frost probability
2. incorporate chill acclimation and deacclimation into our models (See Al Kovaleski's work; ask him for ideas, perhaps)
3. add a column for dormancy duration estimates for preceding winter (based on temperature alone, I suppose)


```{r libraries}
library(ggplot2)
library(dplyr)

```


#Data
```{r}
load(file="/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/phenAbio.RData")

```

# Spring frost index
Calculate the records collected in years following different snow/temperature regimes 


## Frost categories

At this point, `df.shift` is the dataframe with the information that we need! I'll keep it like this for now, but eventually (e.g. after I add the frost index) I'll rename it to `phen.abio` and the rest of the modeling scripts will work. 

```{r}
#check cutoffs
quantile(phen.abio$winter_seas.total.precip, na.rm=TRUE)
quantile(phen.abio$spring_seas.min.temp, na.rm=TRUE)


#based on previous WINTER temperature and precipitation
phen.abio <- phen.abio %>%
  filter(!is.na(winter_seas.total.precip) & !is.na(spring_seas.min.temp)) %>%
  mutate(
    springFrost = case_when(
      winter_seas.total.precip <= quantile(winter_seas.total.precip, 0.25) & spring_seas.min.temp <= quantile(spring_seas.min.temp, 0.25) ~ "high",
      winter_seas.total.precip >= quantile(winter_seas.total.precip, 0.75) & spring_seas.min.temp >= quantile(spring_seas.min.temp, 0.75) ~ "low",
      TRUE ~ "moderate"
    )
  )
table(phen.abio$springFrost)


#Use df.springFrost for the abiotic+phenology dataframe (for models, etc)
phen.abio=df.springFrost
save(phen.abio, file="/Users/elizabethlombardi/Desktop/Research/UNM/Erin phenology project/Phenology collections project/phenAbio.RData")

```

# Plot Data

```{r}
#basic plot with all data for all species in one graph
ggplot(phen.abio, aes(x = year, y = ordinal_date, color = factor(springFrost))) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, aes(group = springFrost)) +  # Add linear regression lines
  scale_color_manual(values = c("red", "goldenrod2", "slategray")) +  # You can adjust colors
  theme_minimal()

#Species plots with linear regression
ggplot(phen.abio, aes(x = year, y = ordinal_date, color=factor(springFrost))) +
  geom_point(outlier.shape = NA) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Box Plot of Values by Category",
    x = "Date",
    y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "top") +
  scale_fill_manual(values=colors) +
  facet_wrap(~scientific_name)


#What if we look at previous winter temperatures instead of spring
#basic plot with all data for all species in one graph
ggplot(df.winter, aes(x = year, y = ordinal_date, color = factor(springFrost))) +
  geom_point() +
  #geom_smooth(method = "lm", se = FALSE, aes(group = springFrost)) +  # Add linear regression lines
  scale_color_manual(values = c("red", "goldenrod2", "slategray")) +  # You can adjust colors
  theme_minimal()
#looks like only one year falls into the low risk category here...
```

Based on these results, it doesn't seem like there is a huge impact of frost index on ordinal date of observation for all data, at least. I'll keep it and re-run PCA with this as a category worth considering. 

It might also be a good idea to add a dormancy variable into the dataset and models. Do do this, I have to get new daily PRISM data for all of the years in the dataset. I would want:
-daily mean temperature?
-daily minimum temperature
-daily maximum temperature
